Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization

### 2014-08-13 TOMS_Paper/assembly.tex
Update autotuner table + minor tweaks

We first comment on the overhead of the autotuning process. In the context of Firedrake, the framework in which COFFEE is integrated, the autotuner is executed at run-time, once the local assembly kernels are provided by the FEniCS Form Compiler. Autotuning, therefore, introduces an overhead in the application execution time. For a given problem instance, such overhead originates from four sources: 1) creation of the various code variants; 2) generation of a C file containing such variants (as simple function calls, plus the ``main'' function that invokes the variants, in sequence); 3) compilation of the autotuning file; 4) execution. Of these four points, we note that: the cost of 4) is relatively small, because each variant's execution time is bound by an empirically-found value (e.g. some milliseconds). The cost of all four points is constrained by our heuristics to prune the search space, as described in Section~\ref{sec:autotune}. Moreover, for a given form and a given discretization, the autotuner needs to be executed only once, since its output is saved and reused for later assemblies. This implies that if the assembly occurs in a time stepping loop, or the same form is executed on a different mesh, or known quantities of the input problem are changed, then the assembly cost is rapidly amortized. Premised that, the most important thing remains that when working with real unstructured meshes - which are likely to be composed of millions of elements, leading to long-lasting assembly phases - the autotuner overhead practically becomes completely negligible. In our experience, and in particular in the four examined problems, the autotuning process lasted less than a minute in the majority of cases. Most often, it took less than thirty seconds. Rarely it needed more than a minute, specifically in the case of hyperelasticity; however, despite the inherent complexity of this form, we measured a peak of only 4 minutes for the extreme case $[q=4, p=4, nf=1]$, while 1.30 minutes were needed in the case $[q=1, p=1, nf=1]$. This analysis certifies that the autotuner overhead is definitely sustainable in real-world applications.

In this paper and in~\cite{Luporini}, we have repeatedly claimed that different forms (and different discretizations) require distinct sets of transformations to reach close-to-peak performance. To demonstrate this, we now report details about the output of the autotuning process. We show that for the four examined forms - more specifically, for the 224 problem instances $[form, q, p, nf]$ illustrated in the previous figures - a plethora of optimization strategies have been selected by the autotuner. We also complement and strengthen our claim by showing the autotuner's output for two additional forms whose performance results, for brevity, were not shown in Section~\ref{sec:perf-results-forms}: a Mass and a Mixed Poisson problems.

Table~\ref{table:autotuner-output} shows the number a given transformation has been selected by the autotuner. To not hinder readability, the output has been grouped by form, rather than showing the selected optimization strategy for each $[form, q, p, nf]$. Values in the \texttt{rewrite strategy} column can either be \texttt{base} or \texttt{aggressive}, as explained in Section~\ref{sec:autotune}: in the former case, only generalized loop-invariant code motion in applied; in the latter case, the rewrite rules are recursively applied as extensive as possible. If the value of the column is \texttt{aggressive} (\texttt{base}) it means that in the majority of cases the \texttt{aggressive} (\texttt{base}) strategy prevailed on the \texttt{base} (\texttt{aggressive}) one. Precomputation of integration-dependent terms was explained in Section~\ref{sec:expr-rewriter-obj}. The \texttt{split} column refers to the expression splitting transformation (\cite{Luporini}, and summary in Section~\ref{sec:summary-opts}). The \texttt{unroll} column indicates the application of explicit unrolling. Other columns are of obvious meaning. Values in the various columns illustrate the number of problem instances out of the total (column \texttt{mumber of variants}) in which an optimization was activated. Obviously, more transformations are typically used in combination in a same problem.

%In the table, we implicitly assume that padding and data alignment (Section~\ref{sec:padding}), as well as precomputation of integration-dependent terms (Section~\ref{sec:expr-rewriter-obj}) have always been applied.

\newcommand{\autotuneroutput}{
\tiny
\begin{tabulary}{1.0\textwidth}{c|c|ccc|ccccc}
\cline{3-10}
\multicolumn{2}{c}{} & \multicolumn{3}{c|}{\texttt{expression rewriting}} & \multicolumn{5}{c}{\texttt{code specialization}} \\
\cline{1-10}
problem & \specialcell{number of\\variants} & \specialcell{rewrite\\strategy} & \specialcell{zero-valued columns\\avoidance} & \specialcell{precompute\\integration terms} & \specialcell{padding\\data alignment} & split & \specialcell{vector-register\\tiling} & unroll & BLAS \\
\cline{1-10}
\texttt{helmholtz} & 64 & aggressive & 0 & 0 & all & 8 & 17 & 15 & 0 \\
\texttt{mass} & 64 & base & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
\texttt{elasticity} & 64 & aggressive & 39 & 0 & all & 0 & 11 & 0 & 2 \\
\texttt{poisson} & 64 & aggressive & 50 & 0 & all & X & X & 1 & 0 \\
\texttt{mixed poisson} & X & base & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
\texttt{hyperelasticity} & 32 & aggressive & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
\cline{1-10}
\end{tabulary}
}

\begin{table*}[t]
\tbl{Autotuner output.}{ \scriptsize \autotuneroutput }
\label{table:autotuner-output}
\end{table*}

>>>>

\paragraph{Autotuning Overhead}
We first comment on the overhead of the autotuning process. In the context of Firedrake, the framework in which COFFEE is integrated, the autotuner is executed at run-time, once the local assembly kernels are provided by the FEniCS Form Compiler. Autotuning, therefore, introduces an overhead in the application execution time. Such overhead originates from four operations: 1) creation of the various code variants; 2) generation of a C file containing such variants (as simple function calls, plus the ``main'' function that invokes the variants, in sequence); 3) compilation of the autotuning file; 4) execution. Of these four points, we note that: the cost of 4) is relatively small, because each variant's execution time is bound by an empirically-found value (e.g. some milliseconds). The cost of all four points is constrained by our heuristics to prune the search space, as described in Section~\ref{sec:autotune}. Moreover, for a given form and a given discretization, the autotuner needs to be executed only once, since its output is saved and reused for later assemblies. This implies that if the assembly occurs in a time stepping loop, or the same form is executed on a different mesh, or known quantities of the input problem are changed, then the assembly cost is rapidly amortized. Premised that, the most important thing remains that when working with real unstructured meshes - which are likely to be composed of millions of elements, leading to long-lasting assembly phases - the autotuner overhead practically becomes completely negligible. In our experience, and in particular in the four examined problems, the autotuning process lasted less than a minute in the majority of cases. Most often, it took less than thirty seconds. Rarely it needed more than a minute, as for instance in the case of hyperelasticity; however, despite the inherent complexity of this form, we measured a peak of only 4 minutes for the extreme case $[q=4, p=4, nf=1]$, while 1.30 minutes were needed in the case $[q=1, p=1, nf=1]$. This analysis certifies that the autotuner overhead is definitely sustainable in real-world applications.

\paragraph{Selected Optimizations}
In this paper and in~\cite{Luporini}, we have repeatedly claimed that different forms (and different discretizations) require distinct sets of transformations to reach close-to-peak performance. To demonstrate this, we now report details about the output of the autotuning process. We show that for the four examined forms - more specifically, for the 224 problem instances $[form, q, p, nf]$ illustrated in the previous figures - a plethora of optimization strategies have been selected by the autotuner. We also complement and strengthen our claim by showing the autotuner's output for two additional forms whose performance results, for brevity, were not shown in Section~\ref{sec:perf-results-forms}: a Mass and a Mixed Poisson problems.

Table~\ref{table:autotuner-output} shows the number a given transformation has been selected by the autotuner. To not hinder readability, the output has been grouped by form, rather than showing the selected optimization strategy for each $[form, q, p, nf]$. Values in the \texttt{rewrite strategy} column can either be \texttt{base} or \texttt{aggressive}, as explained in Section~\ref{sec:autotune}: in the former case, only generalized loop-invariant code motion in applied; in the latter case, the rewrite rules are recursively applied as extensive as possible. If the value is \texttt{aggressive} (\texttt{base}) it means that in the majority of cases the \texttt{aggressive} (\texttt{base}) strategy prevailed on the \texttt{base} (\texttt{aggressive}) one. Precomputation of integration-dependent terms was explained in Section~\ref{sec:expr-rewriter-obj}. The \texttt{split} column refers to the expression splitting transformation (\cite{Luporini}, and summary in Section~\ref{sec:summary-opts}). The \texttt{unroll} column indicates the application of explicit unrolling. Other columns are of obvious meaning. Values in the various cells illustrate the number of problem instances (out of the total reported in column \texttt{number of variants}) in which an optimization was activated. Obviously, more transformations are typically used in combination in a same problem.

\newcommand{\autotuneroutput}{
\tiny
\begin{tabulary}{1.0\textwidth}{c|c|ccc|ccccc}
\cline{3-10}
\multicolumn{2}{c}{} & \multicolumn{3}{c|}{\texttt{expression rewriting}} & \multicolumn{5}{c}{\texttt{code specialization}} \\
\cline{1-10}
problem & \specialcell{number of\\variants} & \specialcell{rewrite\\strategy} & \specialcell{zero-valued columns\\avoidance} & \specialcell{precompute\\integration terms} & \specialcell{padding\\data alignment} & split & \specialcell{vector-register\\tiling} & unroll & M \\
\cline{1-10}
\texttt{helmholtz} & 64 & aggressive & 0 & 0 & all & 12 & 26 & 20 & 0 \\
\texttt{mass} & 64 & base & 0 & 0 & all & 0 & 0 & 31 & 2 \\
\texttt{elasticity} & 64 & aggressive & 39 & 0 & all & 0 & 11 & 0 & 2 \\
\texttt{poisson} & 64 & aggressive & 52 & 0 & all & all & 0 & 0 & 0 \\
\texttt{mixed poisson} & 192 & base & 0 & 0 & all & 32 & 68 & 36 & 0 \\
\texttt{hyperelasticity} & 32 & aggressive & all & all & all & 2 & 0 & 0 & 0 \\
\cline{1-10}
\end{tabulary}
}

\begin{table*}[t]
\tbl{Summary of the optimizations selected by the autotuner in a number of forms}{ \scriptsize \autotuneroutput }
\label{table:autotuner-output}
\end{table*}
